Transmembrane helices are also an overlooked source of major histocompatibility complex class II epitopes and evolutionary more conserved than expected by chance

### 2020-09-23 bbbq_article.tex
Process feedback @bogeert, progress https://github.com/richelbilderbeek/bbbq_article/issues/69

\paragraph{Classification of HLA}

Any human's immune system detects only a fraction of all possible
peptide fragments.
For humans, the MHC proteins are encoded by the
HLA (Human Leukocyte Antigens) genes.
Each MHC complex can only bind a subset of all possible peptides.
For example, HLA-A and HLA-B have no overlap in which
peptides they bind (\cite{lund2004definition}).
The HLA region of humans is highly polymorphic, 
making it hard to classify all of the many haplotypes.

\paragraph{Epitope prediction}

@@

\paragraph{TMHs}

Transmembrane helices are structurally conserved structures that span
a cell membrane with an alpha helix.
TMHs are hydrophobic, as this is required to span the 
hydrophobic cellular lipid membrane. Additionally,
they often have a length of 23 amino acids to be able to span
the membrane.
Polypeptide fragments derived from TMHs are among the most hydrophobic,
together with the internals of soluble proteins, where the
hydrophobicity parts guide the protein to achieve its 3D configuration.
TMHs are general structures: 25 percent of the human proteome is
anchored by at least one TMH. SARS-CoV-2 has 21 TMHs, 
making up for 5.5 percent of its entire genome.
There are multiple computational tools developed to predict which
parts of membrane proteins are TMH.

\paragraph{MHC-I presents hydrophobic regions more often}

One might expect that the hydrophobic epitopes presented,
for a same hydrophobicity, are as likely to stem from 
membrane proteins TMHs or from soluble proteins hydrophobic (non-TMH) regions.
For MHC-I, however, it is found that the 9-mers stemming
from TMHs are presented more often than expected by 
chance (\cite{bianchi2017})
\richel{but also when compensating for hydrophobicity?}.
For MHC-II, it is unknown which percentage of binders 
is derived from TMHs for all 13-mers present in a proteome
that have the same hydrophobicity.

\paragraph{HLAs increase detection range}

One of many strategies to increase the number of pathogenic
peptides that can be detected is to express a wide variety of
MHCs, which is assured by the human leukocyte antigen gene complex (HLA).
However, a different HLA will result in a different variety of MHCs,
that will display different peptide fragments. Additionally, there
may have been selection the HLA to display and or detect different
peptide fragments.

\paragraph{MHC-I presents TMH-derived epitopes in humans more often}

For MHC-I, it was found that predicted epitopes derived 
from human transmembrane helices (TMHs)
are over-presented by all 5 HLA-A and 
most of 8 HLA-B super types (\cite{bianchi2017}).
One explanation is that the presentation of TMHs 
may have an evolutionary advantage for 
the (human) host, as TMHs have a reduced variability 
due to the functional requirement of being able to span a lipid bilayer. 
Due to this, pathogens have a lower chance to develop an escape mutation,
as many mutations will result in a dysfunctional TMH.
Note that the mechanism by which a cell presents its TMHs is
yet unknown.

>>>>

% \paragraph{Classification of HLA}

Any human's immune system detects only a fraction of all possible
peptide fragments.
For humans, the MHC proteins are encoded by the
HLA (Human Leukocyte Antigens) genes.
There are three genes encoding for MHC-I, which are HLA-A, HLA-B and HLA-C,
for MHC-II there are three major genes, which are HLA-DR, HLA-DQ, HLA-DP.
Each MHC complex can only bind a subset of all possible peptides.
For example, HLA-A and HLA-B have no overlap in which
peptides they bind (\cite{lund2004definition}).
The HLA region of humans is highly polymorphic, with hundreds
to thousands of alleles for a gene (\cite{marsh2010nomenclature}).

% \paragraph{HLAs increase detection range}

Because of these multiple and highly polymorphic MHC genes,
the number of pathogenic peptides that can be detected is increased,
as a wide variety of MHCs will be expressed,
each presenting their own subset of epitopes.

% \paragraph{Epitope prediction}

@@

% \paragraph{TMHs}

As the immunoproteasome is a cytosolic protein complex, on expects it
to cleave only cytosolic proteins. However, it has recently been
discovered that transmembrane protein residues are also presented
by MHC-I complexes (\cite{bianchi2017}). This discovery is based
on both in vitro elution studies, as well as on in silico predictions,
with a focus on transmembrane helices. 
The intracellular pathway, in which transmembrane residues are loaded onto
an MHC, however, is yet unknown.

Transmembrane helices are structurally conserved structures that span
a cell membrane with an alpha helix.
TMHs are hydrophobic, as this is required to span the 
hydrophobic cellular lipid membrane. Additionally,
they often have a length of 23 amino acids to be able to span
the membrane.
Polypeptide fragments derived from TMHs are among the most hydrophobic,
together with the internals of soluble proteins, where the
hydrophobicity parts guide the protein to achieve its 3D configuration.
TMHs are general structures: 25 percent of the human proteome is
anchored by at least one TMH. Also the pathogen SARS-CoV-2 has 21 TMHs, 
making up for 5.5 percent of its entire genome.
There are multiple computational tools developed to predict which
parts of membrane proteins are TMH (\cite{krogh2001predicting,bianchi2017,kall2004combined,arai2004conpred,jones2007improving,klammer2009metatm,wang2019efficient}).

% \paragraph{MHC-I presents TMH-derived epitopes in humans more often}

For MHC-I, it was found that predicted epitopes derived 
from human transmembrane helices (TMHs)
are over-presented by all 5 HLA-A and 
most of 8 HLA-B super-types (\cite{bianchi2017}),
due the high hydrophobicity as is required for a TMH to span
the (hydrophobic) cell membrane, combined with the fact that the MHC
binding cleft is hydrophobic as well.

It is unknown if the presentation of hydrophobic/TMH residues
is evolutionary selected for.
One possible explanation is that TMHs have a reduced variability,
imposed by the functional requirement of being able to span a lipid bilayer.
Due to this, the TMHs that pathogens possess, 
have a lower chance to develop an escape mutation,
as many mutations will result in a dysfunctional TMH.
